---
title: "CHJ Day 4 Exercises"
output: html_document
---

```{r importing_data_real, include=FALSE, echo=FALSE}
library(tidyverse)
library(learnr)
library(lubridate)
library(readxl)
library(DT)
library(flair)
library(janitor)

sd <- read_csv("san_diego.csv")

sd <- clean_names(sd)


sd_adjusted <- sd %>% 
  mutate(death_date=mdy(death_date)) %>% 
  mutate(month=month(death_date, label=TRUE, abbr=TRUE)) 

sd_month <- sd_adjusted %>% 
  count(year, month, manner_of_death, name="deaths") %>% 
  mutate(date=mdy(paste0(month, " 1, ", year)))

scented <- read_excel("Scented_all.xlsx", sheet=1)
unscented <- read_excel("Unscented_all.xlsx", sheet=1)

scented <- scented %>% 
  mutate(type="scented")
  
unscented <- unscented %>% 
  mutate(type="unscented")

scented <- bind_rows(scented, unscented)

scented_summary <- scented %>%
  arrange(Date) %>%
  filter(Date >= "2017-01-01") %>%
  mutate(Date=ymd(Date)) %>% 
  filter(CandleID <= 3) %>%
  group_by(Date, type) %>%
  summarise(Average_Rating=mean(Rating),
            Reviews=n())


no_scent_analysis <- scented %>% 
  # filter the reviews so we're only looking at scented candles
  filter(type=="scented") %>% 
  arrange(Date) %>% 
  # only focus on the reviews from this year
  filter(Date >= mdy("01-01-2020")) %>% 
  # create a new column no_scent that will fill a 1 if No Scent is mentioned
  mutate(no_scent=case_when(
    str_detect(Review, "[Nn]o scent") ~ 1, 
    str_detect(Review, "[Nn]o smell") ~ 1,
    str_detect(Review, "[Dd]oes not smell like") ~ 1,
    str_detect(Review, "[Dd]oesn't smell like") ~ 1,
    str_detect(Review, "[Cc]an't smell") ~ 1,
    str_detect(Review, "[Cc]annot smell") ~ 1,
    str_detect(Review, "[Ff]aint smell") ~ 1,
    str_detect(Review, "[Ff]aint scent") ~ 1,
    str_detect(Review, "[Dd]on't smell") ~ 1,
    str_detect(Review, "[Ll]ike nothing") ~ 1,
    TRUE ~ 0
  )) %>% 
  # create a new month column
  mutate(month=month(Date, label=TRUE)) %>% 
  # let's aggregate around month
  group_by(month) %>% 
  # we just need to add up the 1s in no_scent and count up the total reviews with n()
  summarize(no_scent=sum(no_scent), reviews=n()) %>% 
  # and do some math
  mutate(percent=no_scent/reviews*100)


```


## Data introduction

Let's take a look at what were working with:

```{r view_unscented}
library(DT)
library(dplyr)

scented <- read_excel("Scented_all.xlsx", sheet=1)
unscented <- read_excel("Unscented_all.xlsx", sheet=1)

scented <- scented %>% 
  mutate(type="scented")
  
unscented <- unscented %>% 
  mutate(type="unscented")

scented <- bind_rows(scented, unscented)

scented %>% 
  head(4) %>% 
  datatable()
```

These are more than 21,000 reviews for the top 5 rated scented and unscented candles on Amazon. It goes back years.

Let's quickly filter this data to the top 3 candles only and figure out the `Average_Rating` for each day. And also count the total `Reviews` for each day. And we should convert the dates in the `Date` column with a **lubridate** function.

### Exercise 1

Fill in the blanks with the right code.

```{r Exercise-1, warning=F, message=F}
scented_summary <- scented %>%
  arrange(Date) %>%
  filter(Date >= "2017-01-01") %>%
  # convert the String Date into a date format using lubridate function
  mutate(Date=___(Date)) %>% 
  filter(CandleID <= 3) %>%
  group_by(Date, type) %>% 
  # Now how do we aggregate average and count?
  _________(Average_Rating=____(Rating),
            Reviews=___)
  
scented_summary
```

We've narrowed down 21,000 rows of data to a more workable 2,471.

This summarized data set will be our entry to the:

## Grammar of Graphics

Look at the difference between these two charts without and with a scales function. 

```{r second_chart}
ggplot(data= scented_summary) +
  geom_point(mapping=aes(x= Date, y= Average_Rating, color=type)) +
  geom_vline(xintercept= ymd("2020-01-20"))
```

Check out the x-axis.

### Exercise 2

Now add `scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month")` to the bottom of the code


```{r Exercise-2, warning=F, message=F}
ggplot(data= scented_summary) +
  geom_point(mapping=aes(x= Date, y= Average_Rating, color=type)) +
  geom_vline(xintercept= ymd("2020-01-20")) +

```

### Exercise 3

Now, try it with nrow=1 or ncol=1

```{r Exercise-3, warning=F, message=F}
ggplot(data= scented_summary) +
  geom_point(mapping=aes(x= Date, y= Average_Rating, color=type)) +
  geom_vline(xintercept= ymd("2020-01-20")) +
  scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
  facet_wrap(vars(type), ___ = 1)
```



### Theme

Change the appearance of anything in the plot.

There are many built-in themes.

* theme_grey()
* theme_bw()
* theme_dark()

### Exercise 4

Try out the different themes listed above in the code below.

```{r exercise-4, tidy=FALSE, message=FALSE}
ggplot(data= scented_summary) +
  geom_point(mapping=aes(x= Date, y= Average_Rating, color=type)) +
  geom_vline(xintercept= ymd("2020-01-20")) +
  scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
  labs(title = "Candle reviews on Amazon.com",
       subtitle = "Top 3 most popular candles",
       x = "Date",
       y = "Average rating",
       color = "Type",
       caption = "Source: Kate Petrova") +
  theme______
```




### Exercise 5

Add this chunk of code in the exercise below it:

```
theme_bw() + 
theme(legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      panel.grid = element_blank(),
      axis.title.y = element_text(face = "italic"))
```

```{r Exercise-5, tidy=FALSE, message=FALSE}
ggplot(data= scented_summary) +
  geom_point(mapping=aes(x= Date, y= Average_Rating, color=type)) +
  geom_vline(xintercept= ymd("2020-01-20")) +
  scale_x_date(date_labels = "%m-%Y", date_breaks = "6 month") +
  labs(title = "Candle reviews on Amazon.com",
       subtitle = "Top 3 most popular candles",
       x = "Date",
       y = "Average rating",
       color = "Type",
       caption = "Source: Kate Petrova") +
  theme______
```


 
```{r no_scent, warning=F, message=F}
 
no_scent_analysis <- scented %>% 
  # filter the reviews so we're only looking at scented candles
  filter(type=="scented") %>% 
  arrange(Date) %>% 
  # only focus on the reviews from this year
  filter(Date >= mdy("01-01-2020")) %>% 
  # create a new column no_scent that will fill a 1 if No Scent is mentioned
  mutate(no_scent=case_when(
    str_detect(Review, "[Nn]o scent") ~ 1, 
    str_detect(Review, "[Nn]o smell") ~ 1,
    str_detect(Review, "[Dd]oes not smell like") ~ 1,
    str_detect(Review, "[Dd]oesn't smell like") ~ 1,
    str_detect(Review, "[Cc]an't smell") ~ 1,
    str_detect(Review, "[Cc]annot smell") ~ 1,
    str_detect(Review, "[Ff]aint smell") ~ 1,
    str_detect(Review, "[Ff]aint scent") ~ 1,
    str_detect(Review, "[Dd]on't smell") ~ 1,
    str_detect(Review, "[Ll]ike nothing") ~ 1,
    TRUE ~ 0
  )) %>% 
  # create a new month column
  mutate(month=month(Date, label=TRUE)) %>% 
  # let's aggregate around month
  group_by(month) %>% 
  # we just need to add up the 1s in no_scent and count up the total reviews with n()
  summarize(no_scent=sum(no_scent), reviews=n()) %>% 
  # and do some math
  mutate(percent=no_scent/reviews*100)

no_scent_analysis
```

**Make me a chart of your choice using the data above**

### Exercise 6
   
```{r exercise-6, tidy=FALSE, message=FALSE}
## whatever you like (okay, I prefer if you use the percent column)
ggplot(no_scent_analysis) +

```


## Putting it all together


```{r sd1, eval=F}
library(tidyverse)
library(janitor)
library(lubridate)

sd <- read_csv("san_diego.csv")

sd <- clean_names(sd)

sd_adjusted <- sd %>% 
  mutate(death_date=mdy(death_date)) %>% 
  mutate(month=month(death_date, label=TRUE, abbr=TRUE)) 

sd_adjusted
```

Let's start summarizing the data so we can start looking for trends.

### Exercise 7

Can you `count` up the number of `deaths` by manner_of_death by month and year, please?

```{r exercise-7, warning=F, message=F}
sd_month <- sd_adjusted %>% 
  _____(year, month, manner_of_death, name="______") %>% 
  # we'll also add a new date column in the proper format for good charting
  mutate(date=mdy(paste0(month, " 1, ", year)))


sd_month
```


### Exercise 8

Now, fill in the blank below to create a line chart for each type of death

```{r exercise-8, fig.width=8, warning=F, message=F}
sd_month %>% 
  ggplot(aes(x=date, y=deaths, color=manner_of_death, group=manner_of_death)) +
  geom_____() +
  labs(title="Monthly deaths in San Diego",
       caption="Source; San Diego Medical Examiner")
```


### By year

I went ahead and created a new `sd_year` dataframe counting up the deaths by year (while excluding October, November, and December) so we can compare prior years to this year.

```{r sd_year_data}
sd_year <- sd_adjusted %>% 
  # if we're going to compare this year to previous years, we need to exclude data we don't have yet
  filter(!month %in% c("Oct", "Nov", "Dec")) %>% 
  count(year, manner_of_death, name="deaths") 

datatable(sd_year)
```

### Exercise 9

Okay, your turn to make a chart.

Make me a faceted chart that breaks out all the individual manner_of_deaths into its own chart, small-multiple style.

```{r exercise-9, fig.width=8}
sd_year %>% ggplot(aes(x=year, y=deaths)) +
  geom_col() +
  ______wrap(vars(_______________), ncol=4)
```


We're going to use a function called `case_when` to create a new column called **year_type**. If the year is 2020, then it will be "2020" otherwise it will be "1997-2020". And then we find the average number of deaths for each month for those two groups.

```{r sd_group, warning=F, message=F}
sd_group <- sd_adjusted %>% 
  filter(!month %in% c("Oct", "Nov", "Dec")) %>% 
  count(year, month, manner_of_death, name="deaths") %>% 
  mutate(year_type=case_when(
    year==2020 ~ "2020",
    TRUE ~ "1997-2019"
  )) %>% 
  group_by(month, manner_of_death, year_type) %>% 
  summarize(avg_deaths=mean(deaths, na.rm=T)) %>% 
  filter(!is.na(month))

datatable(sd_group)
```


Let's chart it.

### Exercise 10

Can you please create a faceted line chart of the data above? But with year_type as two different lines?

Fill in the three blanks to generate the chart.

```{r exercise-10,  fig.width=8, fig.height=5, warning=F, message=F}
ggplot(sd_group, aes(x=_____, y=__________, color=year_type, group=year_type)) +
  geom_____() +
  facet_wrap(vars(manner_of_death), scales="free_y", ncol=2)
```

Let's compare that overall instead of by month.

Here's the code. We're also excluding September based on what we discovered in the chart above.

```{r sd_group_compare}
sd_group_compare <- sd_adjusted %>% 
  filter(!month %in% c("Sep", "Oct", "Nov", "Dec")) %>% 
  count(year, manner_of_death, name="deaths") %>% 
  mutate(year_type=case_when(
    year==2020 ~ "2020",
    TRUE ~ "1997-2019"
  )) %>% 
  group_by(manner_of_death, year_type) %>% 
  summarize(avg_deaths=round(mean(deaths, na.rm=T)))

datatable(sd_group_compare)
```


### Exercise 11

Run the code below with `manner_of_death` as x and `avg_deaths` as y.

Then swap them.

Which do you prefer and why?

```{r exercise-11, fig.width=8}
ggplot(sd_group_compare, aes(x=________, y=______, fill=year_type)) +
  geom_bar(position="dodge", stat="identity") 
```



I want to get rid of the manners of death that have barely any and I want to reorder the labels so that it's in alphabetical order.

### Exercise 12

Take a look at the code below. Absorb it.

Then generate the code and see what pops up.

```{r exercise-12}
sd_group_compare %>% 
  filter(!manner_of_death %in% c("Other", "Family Paid Autopsy")) %>% 
  filter(!is.na(manner_of_death)) %>% 
  ggplot(aes(x=avg_deaths, y=forcats::fct_rev(manner_of_death),  fill=year_type)) +
  geom_bar(position="dodge", stat="identity") +
  labs(title="Manner of death in San Diego",
       subtitle="January and August deaths in 2020 compared to average deaths between 1997 and 2019",
       caption="Source: San Diego Medical Examiner",
       y="Manner of death",
       x="Average deaths",
       fill="Year") +
  theme_minimal()

```
