---
title: "Austin Animal Center Intakes Analysis"
output: html_notebook
---

## Goals of this notebook

We are looking at [Austin Animal Center Intakes](https://data.austintexas.gov/Health-and-Community-Services/Austin-Animal-Center-Intakes/wter-evkm) data, which is animals brought into the official city shelter.

Here are the questions we'll answer:

- Are animal intakes increasing or decreasing each year since 2016?
- Are there seasonal monthly trends in overall animal intakes?
- Do certain types of animals drive seasonal trends?

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
```

## Import the data

> Will need to change this to github version.

We read in the data from a saved csv file. Could also do this with the [RSocrata](https://github.com/Chicago/RSocrata) package.

```{r }
raw_data <- read_csv("data-raw/Austin_Animal_Center_Intakes.csv") %>% clean_names()

# peek at table
raw_data %>% head()
```

### Glimpse to see all at once

```{r}
raw_data %>% glimpse()
```

## Fix a date

This takes a bit to fix the date. An example looks like this;

`01/03/2019 04:19:00 PM`

I need to change this to a real date. I do this by first converting the text to a real date/time value, then extract the date.

```{r}
date_fix <- raw_data %>% 
  mutate(
    intake_date = mdy_hms(date_time) %>% date()
  )

date_fix %>% glimpse()
```

## Create some date helpers

- Now that we have dates, we can parse them out to help us later.
- Reselect dates to put them up front and remove old date values.

```{r}
date_moyr <-date_fix %>% 
  mutate(
    yr = year(intake_date) %>% as.character(), # as char to avoid continuous numb
    mo = month(intake_date, label = TRUE),
  ) %>% 
  select(intake_date, yr, mo, everything(), -date_time, -month_year)

# peek as table
date_moyr %>% head()
```

## Filter to 5+ years of data

We only want the most recent full five years, plus the most recent full month.

First let's check what we have with `summary()`.

```{r}
date_moyr %>% summary()
```

### Filter by date

Based on this, let's do this:

- Filter for later than 2015
- Filter for earlier than Oct 2021
- This is our last data cleaning, so we'll save as `intakes`.

```{r}
intakes <- date_moyr %>% 
  filter(
    intake_date > "2015-12-31",
    intake_date < "2021-10-01"
  )

# check the result
intakes %>% summary()
```

## Animal intakes by year

**Are animal intakes increasing or decreasing each year since 2016?**

We need to count rows based on the year of the record. We need to remove 2021 because it is not a full year.

Then we can chart it.

```{r yr-data}
intakes_yr_data <- intakes %>% 
  count(yr, name = "count_animals") %>% 
  filter(yr < 2021)

intakes_yr_data
```

Plot the data

```{r yr-plot}
intakes_yr_data %>% 
  ggplot(aes(x = yr, y = count_animals)) +
  geom_col() +
  geom_text(aes(label = count_animals), vjust = 2, color = "white") +
  labs(
    title = "Animal intakes drop in COVID era",
    subtitle = str_wrap("The pandemic forced the Austin Animal Center to close the center\nand then change their intake process, causing a drop in animals", 80),
    caption = "Source: Austin Animal Center",
    x = "Year",
    y = "Number of animals"
  )
```

## Intakes by month, split by year

**Are there seasonal monthly trends in overall animal intakes?**

Here we count by both the month name and year.

```{r mo-yr-data}
mo_yr_data <- intakes %>% 
  count(mo, yr, name = "count_animals")

mo_yr_data
```

```{r mo-yr-plot}
mo_yr_data %>% 
  ggplot(aes(x = mo, y = count_animals, group = yr)) +
  geom_line(aes(color = yr)) +
  labs(
    title = "Austin Animal Center intakes by month, 2016-2021",
    color = "Year",
    x = "Month",
    y = "Number of animals",
    caption = "Source: Austin Animal Center"
  )
```

With this we can see a spike in spring months, especially in pre-pandemic data. Let's figure out why.

## Seasonal trends by animal

**Do certain types of animals drive seasonal trends?**

Before we dive into this, let's explore what we have for animal types.

```{r chk-type}
intakes %>% 
  count(animal_type)
```

Dogs and cats account for most of the records. Let's recode the data to combine the "Bird" and "Livestock" values into "Other".

### Recode Bird, Livestock as Other

We use `recode()` to change the values. I'm OK overwriting `animal_type` because I'm creating a new R object as I do so. I can always go back to the `intakes` data if needed.

```{r recode-type}
intake_types <- intakes %>% 
  # recode some values as other
  mutate(
    animal_type = recode(animal_type,
      "Bird" = "Other",
      "Livestock" = "Other"
    )
  )

# checking the result
intake_types %>% 
  count(animal_type)
```

### Summarize data by month and animal type

```{r type-data}
intake_type_data <- intake_types %>% 
  filter(
    yr < 2021
  ) %>% count(mo, animal_type, name = "count_animals")

intake_type_data
```

Plot the results

```{r type-plot}
intake_type_data %>% 
  ggplot(aes(x = mo, y = count_animals, group = animal_type)) +
  geom_point(aes(color = animal_type)) +
  geom_line(aes(color = animal_type)) +
  labs(
    title = "A good title, source and everything",
    color = "Animal type",
    x = "Month",
    y = "Number of animals"
  )

```

## Let's try an average for fun

The trend should be the same, but let's figure out how to get the "average" number of intakes by animal type in case there is a single year that affects all years.

```{r}
intake_types %>% 
  filter(yr < 2021) %>% 
  count(mo, yr, animal_type, name = "count_animals") %>% 
  group_by(mo, animal_type) %>% 
  summarize(
    avg_animals = mean(count_animals) %>% round()
  ) %>% 
  ggplot(aes(x = mo, y = avg_animals, group = animal_type)) +
  geom_line(aes(color = animal_type)) +
  geom_point() +
  theme_light() +
  labs(
    title = "Average monthly animal intakes, 2015-2020",
    x = "Month",
    y = "Average animals brought in",
    color = "Animal type"
  )
```

