---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(learnr)
library(fs)
knitr::opts_chunk$set(echo = FALSE)
```


## Topics of the day

## Binding rows (stacking data)

```{r prep-names}
yob2017 <- read_csv(
  "yob2017.txt",
  col_types = cols(sex = col_character()
  )
)

yob2018 <- read_csv(
  "yob2018.txt",
  col_types = cols(sex = col_character()
  )
)

yob2019 <- read_csv(
  "yob2019.txt",
  col_types = cols(sex = col_character()
  )
)

```

We assume the multiple data frames have aleady been imported.

```{r bind-rows, exercise=T, exercise.setup="prep-names"}
# Available data frames: yob2017, yob2019, yob2019

# peek at yob2017. All dfs have same col names.
yob2017 %>% head()

# length of yob2017
yob2017 %>% nrow()

names <- bind_rows(
  "2017" = yob2017,
  "2018" = yob2018,
  "2019" = yob2019,
  .id="yr"
)

names %>% nrow()
```



- bind_rows (with `.id=id` and `"group 1" = one, "group 2" = two, .id = "groups"`)

this might allow stacking dfs by using df's name?

```r
bind_rows(list_of_DFs, .id = "id") %>% 
  spread(key = "id", value = "result")
```

## Joining data frames



### Andrew

- Reshaping data with tidyr.
- Introduction to ggplot.


## Examples

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

