---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Topics of the day

Continuing data wrangling with dplyr. Creating new columns, cleaning data, etc. Datatables.

- select
- filter
- mutate
- group_by and summarize

## Stuff to consider

Import NYTimes covid case numbers


```{r}
covid <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
```

Get daily cases: Since cases and deaths are cumulative, this subtracts the previous day to get the daily count.

```{r}
covid_daily <- covid %>%
  group_by(county, state, fips) %>%
  mutate(
    daily_cases = case_when(
      is.na(lag(cases)) ~ cases,
      TRUE ~ cases - lag(cases)
    ),
    daily_deaths = case_when(
      is.na(lag(deaths))~ deaths,
      TRUE ~ deaths-lag(deaths)
    )
  )
```


```{r}
covid_daily %>% 
  filter(state == "Texas", county == "Travis")
```


Covidtracking.co api

[found here](https://covidtracking.com/data/api)

```{r}
ctapi <- read_csv("https://api.covidtracking.com/v1/states/current.csv")
```


## Examples

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

