---
title: "Coronavirus racial disparities"
author: "Andrew Ba Tran"
output: html_document
---


What happens when you don't have race data? But you have data by county?

Well, you could use demographic data from the Census as a proxy.

Let's load up some libraries.

```{r setup, warning=F, message=F}
# Loading libraries
library(tidyverse)
library(tidycensus)
library(viridis)
library(knitr)
library(readxl)
library(DT)
# Sub in your Census API key here
census_api_key("YOURKEYGOESHERE")
```

Let's import some Covid-19 data. 

```{r covid_import, warning=F, message=F}
# From... The New York Times. Don't tell anyone on me.

corona <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

corona
```

Looks like we have data by county and date.

Let's bring in race population data by county from the Census with our new fancy API function from **tidycensus**..

```{r census_import}
# Downloading race table from the Census and calculating white percent of population
county_race <- get_acs(geography = "county",
                       variables = c("B03002_003", "B03002_004", "B03002_006", "B03002_012", "B03002_007"),
                       summary_var = "B03002_001",
                       geometry = TRUE,
                       shift_geo = TRUE) %>% 
  mutate(pct = round(100 * (estimate/summary_est),2)) %>% 
  mutate(race=case_when(
    variable=="B03002_003" ~"White",
    variable=="B03002_004" ~"Black",
    variable=="B03002_006" ~"Asian",
    variable=="B03002_012" ~"Hispanic",
    variable=="B03002_007" ~"American Indian",
    TRUE ~ "Other"
  ))

county_race_wide <- county_race %>% 
  ungroup() %>% 
  select(GEOID, summary_est, pct, race) 

# let's get rid of the geometry column for now...
county_race_wide$geometry <- NULL

county_race_wide
```

Okay, we have the **corona** table, which has county cases and deaths by day.

And we have county_race_wide data, which we used to calculate the percent of race population in each county.

We need to filter these two data sets:

* corona: We only want the most recent date of data

* county_race_wide: We only want the row with the plurality race

```{r simplify}
# Filtering by the latest dated entry for the most up-to-date figures by county
corona_latest <- corona %>% 
  group_by(fips) %>% 
  arrange(fips, date) %>% 
  slice(n())

# Filtering race county data so the plurality race is the only row per county
county_race_wide <- county_race_wide %>% 
  group_by(GEOID) %>% 
  arrange(desc(pct)) %>% 
  slice(1)

county_race_wide
```

```{r prep_join}
# prepping county race 
# joining the county results data to the corona data
# calculating per capita cases and deaths for each county
corona_race_join_save <- corona_latest %>% 
  left_join(county_race_wide, by=c("fips"="GEOID")) %>% 
  mutate(cases_per_capita=cases/summary_est*100000,
         deaths_per_capita=deaths/summary_est*100000) %>% 
  filter(!is.na(race))

# Calculating median and average cases/deaths per capita by race group
# This excludes counties with no cases and no deaths
corona_race_table <- corona_latest %>% 
  left_join(county_race_wide, by=c("fips"="GEOID")) %>% 
  group_by(race) %>%
  filter(!is.na(race)) %>% 
  summarize(counties=n(),
            population=sum(summary_est),
            cases=sum(cases, na.rm=T),
            deaths=sum(deaths, na.rm=T),
            cases_per_capita=round(cases/population*100000,2),
            deaths_per_capita=round(deaths/population*100000,2))

corona_race_table
```

```{r nchs, warning=F, message=F}

nchs <- read_excel("data/NCHSURCodes2013.xlsx")

nchs <- nchs %>% 
  mutate(`FIPS code` = as.character(`FIPS code`),
         `FIPS code` = case_when(
           nchar(`FIPS code`)==4 ~ paste0("0",`FIPS code`),
           TRUE ~ `FIPS code`
         )) %>% 
  mutate(county_category=case_when(
    `2013 code`==1 ~ "Large central metro",
    `2013 code`==2 ~ "Large fringe metro",
    `2013 code`==3 ~ "Medium metro",
    `2013 code`==4 ~ "Small metro",
    `2013 code`==5 ~ "Micropolitan",
    `2013 code`==6 ~ "Non-core"
  ))

# prepping county race 
# joining the county results data to the 
corona_race_join <- corona_latest %>% 
  left_join(county_race_wide, by=c("fips"="GEOID")) %>% 
  select(-pct, -race) %>% 
  left_join(nchs, by=c("fips"="FIPS code")) %>% 
  # calculating minority percent by subtracting percent white from 100
  mutate(cases_per_capita=cases/summary_est*100000,
         deaths_per_capita=deaths/summary_est*100000#,
         ) %>% 
  filter(!is.na(cases_per_capita))

# Calculating median and average cases/deaths per capita by county group
# This excludes counties with no cases and no deaths

corona_nchs_table <- corona_race_join %>% 
  rename(code=`2013 code`) %>% 
  group_by(code, county_category) %>% 
  summarize(counties=n(),
            population=sum(summary_est),
            cases=sum(cases, na.rm=T),
            deaths=sum(deaths, na.rm=T),
            cases_per_capita=round(cases/population*100000,2),
            deaths_per_capita=round(deaths/population*100000,2)) %>% 
    select(-population)

corona_nchs_table
```

## Look up cases and deaths by county

Use the filter field and sort buttons, please.

```{r counties_drilldown}
corona_race_join_save %>% 
  ungroup() %>% 
  mutate(cases_per_capita=round(cases_per_capita,2),
         deaths_per_capita=round(deaths_per_capita,2)) %>% 
  select(state, county, majority=race, pct, cases, deaths, population=summary_est, `Cases per 100k`=cases_per_capita, `Deaths per 100k`=deaths_per_capita) %>%
  arrange(desc(`Deaths per 100k`)) %>% 
  datatable(filter='top')
```


